---
title: "Tutorial 2: OLS and first repliction"
format: html
editor: visual
toc: true
toc-title: ""
toc-location: right
---

```{r, echo=F,eval=T,message=F,warning=F}
packages <- c("ggplot2", "readxl", "dplyr", "tidyr","xtable")
# Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}
# Packages loading
invisible(lapply(packages, library, character.only = TRUE))
```

Last week, we learnt how to manage simple datasets using R. We are going replicate our very first paper!

## Mankiw, Romer, Weil (1994)

This paper is ...

### Main objectives

1.  Open the dataset
2.  Inspect and clean
3.  Run the main regression
4.  Make a graph

### Data source

Download the data from Moodle.

### Folder organization

We are going to work in `tutorial2`. You may want to create additional subfolders like `raw_data`, `code`, `output` to store the raw data, the code, and the output we create.

``` plaintext
/Users/mmoglia/Dropbox/
└── courses/
    └── 2025_eco1s002
      └── tutorial2
```

```{r}
path <- "/Users/mmoglia/Dropbox/courses/polytechnique/2025_eco1s002/tutorial2"
```
